# Data in geoChronR {#inputs}

geoChronR was designed for data structured in the [LiPD format](http://lipd.net). 
Actually, much of the LiPD framework was developed to accommodate the needs of geoChronR, so by far, you'll find it easiest to work with data in the structured, metadata-rich LiPD format. 
All of the examples and exercises in this book will use LiPD datasets. 
If you have data you want to analyze in geoChronR, the best and easiest way to get those data into the LiPD format is at the [LiPD playground](http://lipd.net/playground).
It takes some time to learn the structure, and some time to annotate your data, but it will save you time down the road. 
If you **really** don't want to lipdify your data, we are working on ways to take advantage of geoChronR tools with simpler data structures. 
This will mean more work in geoChronR, since you'll have to specify a lot information that would normally be included in a LiPD file, but there are times when this makes sense. 
Once this functionality is operable, we'll add a module demonstrating how this works. 

## Loading a LiPD file

The `lipdR` package is your gateway to working with LiPD data. To install `lipdR`, you'll need the remotes package if you don't already have it.

```{r, echo=TRUE, eval=FALSE}
install.packages("remotes")
remotes::install_github("nickmckay/lipdR")
```

Now you can load the lipdR library, and load some data. 

```{r}
library(lipdR)
L <- readLipd("https://lipdverse.org/Temp12k/current_version/GEOFAR_KF16_MgCa.Repschlager.2016.lpd")
```

You can load a LiPD file from the web, as shown above, or locally on your computer. If you leave the input blank (e.g. `L <- readLipd()`), you can interactively choose a file on your computer. 


:::: {.blackbox data-latex=""}
::: {.exercise name="Loading LiPD data"}

Explore [the Temp12k collection on lipdverse.org](https://lipdverse.org/Temp12k/current_version), and load a file into R using lipdR with two different approaches: a) directly using the url as above, and b) download the file to your computer, and load it interactively. 
:::
::::
<details>
  <summary>Click for Answer to 1.1a</summary>
   We'll use [this dataset](https://lipdverse.org/Temp12k/current_version/HeshangCave.Wang.2018.html).
   
   a. First we'll get it directly from the web like this:

```{r, echo=TRUE, eval=TRUE}
    lipd <- readLipd("https://lipdverse.org/Temp12k/current_version/HeshangCave.Wang.2018.lpd")
```

note that the extension is ".lpd" not ".html"
</details> 

<details>
  <summary>Click for Answer to 1.1b</summary>
   b. This time, just go to [the dataset](https://lipdverse.org/Temp12k/current_version/HeshangCave.Wang.2018.html) and download the data to your computer, then run 

```{r, echo=FALSE, eval=FALSE}
    lipd <- readLipd()
```

and select the file interactively. 
</details> 


## Loading multiple LiPD datasets



