# Installing the software you need {#installation}

You need [R (>=3.6)](https://cran.r-project.org/) and we strongly recommend [RStudio as an IDE](https://www.rstudio.com/products/rstudio/download/). 

If you're not sure which version of R you're running, try 

```{r}
R.version
```

If that gives you an error, you likely have an older version, but this should work.

```{r}
sessionInfo()
```


## Installing lipdR

The `lipdR` package is your gateway to working with LiPD data. To install `lipdR`, you'll need the remotes package if you don't already have it.

```{r, echo=TRUE, eval=TRUE}
install.packages("remotes")
remotes::install_github("nickmckay/lipdR")
```


## Installing geoChronR

Finally, you can install geoChronR using a similar command:
```{r, echo=TRUE, eval=FALSE}
remotes::install_github("nickmckay/GeoChronR")
```


This should also install all the dependencies you need. geoChronR is built on top of a large collection of other packages, which means it can take a while to install, but also it means that changes to those packages sometimes causes errors in geoChronR. If you get an error during installation, [please report it here]("http://github.com/nickmckay/GeoChronR/issues/"), and we'll fix it as soon as possible. 






